---
title: "Final Project"
author: "Tsega Abebe"
date: "`r Sys.Date()`"
output: html_document
---

```{r}
# Libraries
library(tidyverse)
library(skimr)
library(caret)
library(dplyr)
```



```{r}
#Load the dataset
cell2cell <- read.csv("cell2celltrain.csv", stringsAsFactors = TRUE)
head(cell2cell)
```

*Data Exploration*
```{r}
#Check the structure
str(cell2cell)
```
We can see that the dataset has 51,047 rows and 58 columns. There are a mix of data types:
- Numeric: 26 columns
- Categorical: 23 columns
- Integer: 9 columns


```{r}
# Check for missing values and summary statistics
skim(cell2cell)
```
- Almost all numeric and catagorical are right skewd 
- I see some missing and unusual values:
  - Negative Revenue: The minimum MonthlyRevenue is -$6.17, which could represent credits or adjustments.
  - Negative TotalRecurringCharge: The minimum is -$11.00, which is unusual for a recurring charge.
  - AgeHH1 and AgeHH2 both have values of 0, which we need to investigate

- I also see extremely high values:
  - Max MonthlyRevenue: $1,223.38 (compared to median of $48.46)
  - Max MonthlyMinutes: 7,359 mins (compared to median of 366)
  - Max DroppedCalls: 221.7 (compared to median of 3)
  - Max CustomerCareCalls: 327.3 (compared to median of 0)


```{r}
#check all the missing values
sapply(cell2cell, function(x) sum(is.na(x)))
```

Some columns have missing values:
- MonthlyRevenue, MonthlyMinutes, TotalRecurringCharge, DirectorAssistedCalls, OverageMinutes, RoamingCalls: 156 missing values each.
- PercChangeMinutes, PercChangeRevenues: 367 missing values
- AgeHH1, AgeHH2: ~909 missing values.
- Handsets, HandsetModels, CurrentEquipmentDays: 1 missing value each.



Lets check the distribution of the churn
```{r}
#Distribution of the churn
ggplot(cell2cell, aes(x = Churn, fill = Churn)) +
  geom_bar() +
  labs(title = "Distribution of Customer Churn", x = "Churn", y = "Count") +
  theme_minimal()
```


```{r}
# Check the proportion of churn cases
table(cell2cell$Churn) %>% prop.table()
```

The data is imbalanced:
- No (Non-Churn): 71.2%
- Yes (Churn): 28.8%



To handle the missing values lets first check from which class the missing values are from:
```{r}
# Check if missing values are mainly from the majority class (non-churned)
cell2cell %>%
  filter(if_any(everything(), is.na)) %>%
  group_by(Churn) %>%
  summarize(count = n())

```

- Since the missing values come from the majority (non-churned) class, it is reasonable to drop these rows. Dropping these rows will have a minimal impact on class representation.



```{r}
# Drop rows with any missing values
cell2cell <- cell2cell %>% drop_na()

# new dimensions
dim(cell2cell)

# check na
sum(is.na(cell2cell))
```



- Let's visualize the distribution and spot any outliers or unusual values. 
- Let's start with variables that are imporatnt to my analysis, like (MonthlyRevenue, MonthlyMinutes, and perhaps AgeHH1/AgeHH2 for customer demographics).

*Monthly Revenue Distribution by Churn*
```{r}
ggplot(cell2cell, aes(x = MonthlyRevenue)) + geom_histogram(bins = 30, fill = "blue", color = "black") + labs(title = "Distribution of Monthly Revenue", x = "Monthly Revenue", y = "Count")
```

- The distribution of Monthly Revenue is highly right-skewed with most customers in the lower revenue brackets.


```{r}
ggplot(cell2cell, aes(x = MonthlyRevenue, fill = Churn)) +
  geom_boxplot(alpha = 0.6) +
  labs(title = "Monthly Revenue Distribution by Churn", x = "Monthly Revenue", y = "Density") +
  theme_minimal()
```
- There are a significant number of outliers as seen in the boxplot, with some revenue values extending well beyond the interquartile range (IQR)

- This indicates that most customers generate low to moderate revenue, while a few generate exceptionally high revenue.


Now lets check call minutes since it is a key indicator of service usage:


```{r}
#Histogram for MonthlyMinutes
ggplot(cell2cell, aes(x = MonthlyMinutes)) + geom_histogram(bins = 30, fill = "orange", color = "black") + labs(title = "Distribution of Monthly Minutes", x = "Monthly Minutes", y = "Count")
```

- Similar to revenue, the distribution is also right-skewed, indicating that most customers have low to moderate usage, while a few have very high usage.


Lets check the box plot for outliers: 
```{r}
ggplot(cell2cell, aes(x = MonthlyMinutes, fill = Churn)) +
  geom_boxplot(alpha = 0.6) +
  labs(title = "Monthly Minutes Distribution by Churn", x = "Monthly Minutes", y = "Density") +
  theme_minimal()

```

- The boxplot shows a high number of outliers, which might indicate irregular usage patterns among a small group of customers.


*Distribution of Household Age (AgeHH1 and AgeHH2)*
```{r}
ggplot(cell2cell, aes(x = AgeHH1)) + geom_histogram(bins = 30, fill = "purple", color = "black") + labs(title = "Distribution of Primary Household Age (AgeHH1)", x = "AgeHH1", y = "Count")

ggplot(cell2cell, aes(x = AgeHH2)) + geom_histogram(bins = 30, fill = "red", color = "black") + labs(title = "Distribution of Secondary Household Age (AgeHH2)", x = "AgeHH2", y = "Count")
```

We need to investigate why why zeros exist and whether they are valid or missing data.

```{r}
# Check proportion of zeros in AgeHH1 and AgeHH2
cell2cell %>%
  summarise(
    AgeHH1_Zeros = sum(AgeHH1 == 0, na.rm = TRUE) / n(),
    AgeHH2_Zeros = sum(AgeHH2 == 0, na.rm = TRUE) / n()
  )
```

- AgeHH1 Zeros: A primary household member aged 0 is biologically impossible. These are likely data entry errors or placeholders for missing values.

- AgeHH2 Zeros: Valid if they represent households with no secondary member (e.g., single-person households).


*Service Quality: Dropped Calls vs Churn*

```{r}
ggplot(cell2cell, aes(x = DroppedCalls, y = Churn, fill = Churn)) +
  geom_boxplot() +
  labs(title = "Dropped Calls vs Churn", x = "Dropped Calls", y = "Churn") +
  theme_minimal()
```


*MonthsInService vs Churn Rate*
Let's see if newer or long-term customers churn more:
```{r}
cell2cell %>%
  group_by(MonthsInService) %>%
  summarise(ChurnRate = mean(Churn == "Yes")) %>%
  ggplot(aes(x = MonthsInService, y = ChurnRate)) +
  geom_line(color = "steelblue") +
  geom_point(color = "steelblue") +
  labs(title = "Churn Rate by Months", x = "Months in Service", y = "Churn Rate") +
  theme_minimal()
```


*Customer Care Calls Impact*
- High call volume often predicts churn:

```{r}
cell2cell %>%
  count(CustomerCareCalls, Churn) %>%
  group_by(CustomerCareCalls) %>%
  mutate(Proportion = n / sum(n)) %>%
  filter(Churn == "Yes") %>%
  ggplot(aes(x = factor(CustomerCareCalls), y = Proportion)) +
  geom_col(fill = "tomato") +
  labs(title = "Churn Rate by Customer Care Calls", x = "Customer Care Calls", y = "Churn Rate") +
  theme_minimal()
```

*Income Group vs Churn*
Check if lower-income customers churn more
```{r}
cell2cell %>%
  group_by(IncomeGroup) %>%
  summarise(ChurnRate = mean(Churn == "Yes", na.rm = TRUE)) %>%
  ggplot(aes(x = factor(IncomeGroup), y = ChurnRate)) +
  geom_col(fill = "darkgreen") +
  labs(title = "Churn Rate by Income Group", x = "Income Group", y = "Churn Rate") +
  theme_minimal()
```



